### 2022-10-13 dissertation/chapters/02-artificial-neural-networks.tex
✨ (disseration): Chapter 2 - Done

\noindent
The human brain is one of the most complicated biological organs in nature. It gives us the ability to think and learn. The field of \ac{ML} has taken a lot of inspiration from the biological brain which lead to the development of \acp{ANN} \cite{ref:rosenblatt:1958}. \acp{ANN} are used throughout this dissertation as the \textit{model} to be trained using \acp{HH}.

The purpose of this chapter is to provide the necessary background information needed on \acp{ANN} and is thus structured into the following sections.

\begin{itemize}
    \item \textbf{Section \ref{sec:anns:bn}} gives background information on the \ac{BN}.

    \item \textbf{Section \ref{sec:anns:an}} introduces the \ac{AN}. Brief discussions follow on input, \index{weights}weights, \index{biases}biases, \index{net input signal}net input signal, \index{activation functions}activation functions, and output.

    \item \textbf{Section \ref{sec:anns:ann}} introduces the \ac{ANN}. Brief discussions follow on \ac{ANN} architecture, topology, and \acp{FFNN}.

    \item \textbf{Section \ref{sec:anns:training}} provides details on the training process, training sets, \index {supervised learning}supervised learning, \index{error function}error functions, performance measures and stopping conditions.

@@

\index{Biological Neuron}
\section{Biological Neuron}\label{sec:anns:bn}

This section introduces the \ac{BN} and provides the necessary background information that show how the \ac{BN} has inspired the development of the \ac{AN}.

The biological neural systems are made up of microscopic nerve cells called neurons \cite{ref:jain:1996}. Figure \ref{fig:biological_neuron} illustrates a single \ac{BN}.

@@

\noindent
The main components of the \ac{BN} include the cell body, \index{dendrites}\textit{dendrites} and the \index{axon}\textit{axon}. \Acp{NN} are formed through the connections between the \index{axons}axons and \index{dendrites}dendrites of various neurons. This is known as \index{synaptogenesis}\textit{synaptogenesis} \cite{ref:huttenlocher:1997}. Such a connection is referred to as a \index{synapse}\textit{synapse}. Communication takes place, through the synapse, by electro-chemical pulse and is often referred to as an \textit{activation} or \textit{action potential}.  Communication signals propagate from the \index{dendrites}dendrites, through the cell body to the axon of a neuron, provoking a signal in the post-synaptic neuron \cite{ref:engelbrecht:2007}. The greater the connection between two neurons, the stronger the communication.  Kennedy \cite{ref:kennedy:2016} defines stronger \index{synapse}synapses as ones that contribute more depolarisation to the neural membrane upon activation than weaker ones. Stronger \index{Synapse}synapses have a higher probability of generating an action potential in the post-synaptic neuron. During activation, the pre-synaptic neuron release neurotransmitters that bind to the post-synaptic neuron \cite{ref:khanacademy:synapse}. The frequent release of these molecules cause the \index{synapse}synapse to grow. Connections that grow over time yield stronger signals (learning), while connections that are weak propagate low intensity signals and vanish over time (forgetting). The ability of synapses to strengthen and weaken over time is known as \index{synaptic plasticity}\textit{synaptic plasticity} \cite{ref:huttenlocher:1997}.

\index{Artificial Neuron}
\section{Artificial Neuron}\label{sec:anns:an}

This section introduces the \ac{AN}. Brief discussion follow on the various components that make up the \ac{AN}. These components include inputs, weights, biases, \index{net input signal}net input signal, and output of the \ac{AN}.

\subsection{Components}\label{sec:anns:components}

The \ac{AN} implements a nonlinear mapping from $\mathbb{R}^{I}$ to
$\mathbb{R}^{T}$, usually in the ranges $[0,1]$ or $[-1,1]$, depending on the
activation function used \cite{ref:engelbrecht:2007} and is given
in Equation \eqref{eq:an_function_mapping} below.

@@

\noindent
In Equation \eqref{eq:an_function_mapping} above, $f_{AN}$ is the mapping function produced by the \ac{AN}, $I$ is the total number of dimensions of the input in real-number space ($\mathbb{R}$), and $T$ is the total number of dimensions of the target (desired output) in real-number space.

The \ac{AN} implements various components and is illustrated in Figure \ref{fig:artificial_neuron}.

@@

\noindent
Each of the components of the \ac{AN} is inspired by some element of the \ac{BN}. Brief descriptions of these are given as follows.

@@

    \item \textbf{Biases:} The biases model a mechanism introduced to influence the strength of the activation (output signal) of the \index{AN}\ac{AN}. The bias is represented by $\theta$, in Figure \ref{fig:artificial_neuron}.

    \item \index{activation function}\textbf{Activation function:} The activation function models the action potential of the \ac{BN}\index{BN} based on the net input signal. The \index{activation function}activation function is represented by $f$, in Figure \ref{fig:artificial_neuron}.

@@

Input signals are $I$-dimensional vectors of numerical values that are obtained either through some environment sensor or from other \acp{AN}. Input signals are often referred to as \index{features}\textit{features} or \index{independent variables}\textit{independent variables} \cite{ref:francis:2001}. Throughout this dissertation, a single input vector is referred to as a pattern. Input data must first be pre-processed before it is presented to the \ac{AN}. Input pre-processing techniques are presented in the following sections.

@@

Input pre-processing improves the training process and contributes to the success of the practical application of the \ac{ANN} \cite{ref:kuzniar:2017}. The primary purpose of input pre-processing is to modify the input data so that it can better match predicted output data. There are many techniques to consider when pre-processing input data. These techniques includes methods to encode data to certain formats, to scale to the correct ranges and to deal with incomplete, invalid or irrelevant data. For the purposes of this dissertation, only \textit{data encoding} and \textit{normalisation} is considered.

\index{Encoding}
\subsubsection{Encoding}\label{sec:anns:an:input:encoding}

For qualitative data such as categories, the class labels have to be converted from textual to numerical representations. In general, class labels are encoded as numerical vectors \cite{ref:srinidhi:2018, ref:brownlee:2017:one-hot}. One such encoding technique is referred to as \index{one-hot encoding}\textit{one-hot encoding}. Harris and Harris \cite{ref:harris:2010} describe \index{One-Hot Encoding}one-hot encoding as a group of vector elements where the classes are represented with a single high element (represented by $1$) and all the others low (represented by $0$). Each element $y_k$, where $k$ is the $k$-th element of the encoded vector $\vec{y}$, uniquely refers to a class. An activation, represented by a value of $1$ at a given index thus represents a class associated with that index.

\index{Normalisation}
\subsubsection{Normalisation}\label{sec:anns:an:input:normalisation}

Numerical data must be normalised and scaled appropriately for the \ac{AN}. In general, input data is normalised to a range that is appropriate for the \index{activation function}activation function being used by the \ac{AN}. Normalisation and scaling yields input data that is comparable to the output produced by the \ac{AN}. For the purposes of this disseration, \index{min-max scaler}min-max scaler \cite{ref:al:2006} and \index{standard score scaler}standard score scaler \cite{ref:jain:2005} are considered.

\index{Min-Max Scalar}
\subsubsection{Min-Max Scaler}\label{sec:anns:an:input:min_max_scaler}

The \index{min-max scaler}min-max scaler, also called \index{unity-based normalisation}\textit{unity-based normalisation}, scales values to the range $[0,1]$. The \index{min-max scaler}min-max scaler is used as a pre-processing technique for target values when the \index{sigmoid}\textit{sigmoid} \index{activation function} activation function is used. The \index{min-max scaler}min-max scaler is given in Equation \eqref{eq:min_max_scaler} below.

@@

\noindent
In Equation \eqref{eq:min_max_scaler} above, $x_{i,p}^{'}$ is the normalised form of $x_{i,p}$, the $i$-th dimension of the input vector $\vec{x}_p$, $p \in \{1,2, \dots, P \}$, where $P$ is the total number of input patterns, $x_{i_{min}}$ and $x_{i_{max}}$ are respectfully the minimum and maximum values of the $i$-th dimension for all input vectors $\vec{x}_p$.

\index{Standard Score Scalar}
\subsubsection{Standard Score Scaler}\label{sec:anns:an:input:standard_score_scaler}

The \index{standard score scaler}standard score scaler, also known as the \index{z-score scaler}\textit{z-score scaler}, scales values by subtracting the mean and scaling to the unit variance of each dimension $i$ for all input patterns $\vec{x}_p, p \in \{1,2, \dots, P\}$. The \index{standard score scaler}standard score scaler is used as a pre-processing technique for target values when the \index{hyperbolic tangent}\textit{hyperbolic tangent} \index{activation function}activation function\index{activation function} is used. The \index{standard score scaler}standard score scaler is given in Equation \eqref{eq:standard_score_scaler} below.

@@

\noindent
In Equation \eqref{eq:standard_score_scaler} above, $\mu_i$ and $\sigma^2_i$ are respectfully the mean and unit variance of the $i$-th dimension of all input vectors $\vec{x}_p$.

@@

The connection strength that synapses in the \acs{BN} have is modeled in the \ac{AN} as weight vectors of numerical values associated with each dimension of the input. Weights can either dampen or strengthen the input by some negative or positive numerical value. Changes in the weight associated with a feature changes the influence that that particular feature has on the predicted output. Finding the correct value for each weight, such that the \ac{AN} yields optimal output, is an optimisation problem \cite{ref:thierens:1993}. Research has shown that weight initialisation plays an important role in the efficient training of \acp{ANN} \cite{ref:thimm:1995}.

\index{Weight Initialisation}
\subsubsection{Weight Initialisation}\label{sec:anns:an:weights:initialisation}

Weight initialisation is the process by which candidate solutions (represented by the weights of the \ac{AN}) to the problem are ``placed'' in the search-space. Weight initialisation influences the speed of convergence, the probability of convergence and the generalisation capabilities of \acp{ANN} \cite{ref:fernandez:2001}. Finding the optimal initialisation values for weights is non-trivial and can be seen as another optimisation problem \cite{ref:de:2016, ref:erdogmus:2003, ref:yam:2000}.

\index{Weight Initialisation}Weight initialisation is dependent on the \index{activation function}activation function used. If weights are initialised as values that are too small, the vanishing gradients problem can occur \cite{ref:hanin:2018}. If weights are initialised as values that are too big, output of the activation function would not be in the active range. This leads to unit saturation, and exploding gradients can occur \cite{ref:hanin:2018, ref:yadav:2018}.

@@

\index{Glorot uniform sampling}Glorot uniform sampling is a specialisation of \index{random uniform sampling}random uniform sampling whereby $\omega = \sqrt{\frac{6}{fanin + fanout}}$ and \index{$fanin$}$fanin$ is the number of input neurons to the weight vector and \index{$fanout$}$fanout$ is the number of output neurons from the weight vector.

\index{Glorot Normal Sampling}
\subsubsection{Glorot Normal Sampling}\label{sec:anns:an:weights:glorot_normal_sampling}

@@

\acp{AN} accumulate the net resulting input signal from all input dimensions into a value called the \index{net input signal}\textit{net input signal}, expressed as $net$. This signal is passed to the \index{activation function}activation function, which provokes an artificial action potential in the \ac{AN}.

A common way by which the \index{net input signal}net input signal is calculated is by means of \acp{SU} \cite{ref:engelbrecht:2007}, which compute the \index{net input signal}net input signal as the weighted sum of all input signals as shown in Equation \eqref{eq:summation_units} below.

@@

\noindent
In Equation \eqref{eq:summation_units} above, $x_{i}$ is the $i$-th dimension of the input vector $\vec{x}$, and $v_{i}$ is the $i$-th dimension of the weight vector $\vec{v}$, associated with the given input dimension.

@@

A bias/threshold term $\theta$ is introduced to help translate the output of the \index{activation function}activation function \cite{ref:benitez:1997}. The value of $\theta$ can be learned during the training process along with the weights of the \ac{ANN}. In order to simplify equations, the input and weight vectors are augmented such that the input and hidden layers have an additional neuron/unit, called the \textit{bias unit} \cite{ref:engelbrecht:2007}. The net input signal can then be rewritten to consider the bias unit, leading to an augmented net input signal.

@@

The augmented net input signal, that includes the bias unit, has the form $net^{'} = net - \theta$. The value of $\theta$ is learned along with the weights such that $\theta = x_{i+1}v_{i+1}$. A constant value $x_{i+1} = -1$ can be used. The net input signal as given in Equation \eqref{eq:summation_units} then changes as is shown in Equation \eqref{eq:augmented_vectors} below.

@@

An \index{activation function}activation function is used to model the action potential of the \index{AN}\ac{AN} \cite{ref:ziv:1994, ref:hodgkin:1952}. The \index{activation function}activation function takes as a parameter, the augmented \index{net input signal}net input signal. When the output produced by the activation function surpasses some threshold value $\tau$, we consider that neuron to have ``fired'' an output signal. \index{activation function}Activation functions thus model \textit{phase shift}. In the context of classification problems, \index{activation function}activation functions form decision boundaries between classes. In the context of regression problems, \index{activation function}activation functions try to approximate some function that maps the input data to some target data.

In general, activation functions produce a non-linear mapping of $\mathbb{R}^{I}$ to the range $[0,1]$ or $[-1,1]$ as shown in Equations \eqref{eq:an_function_mapping_0_1} and \eqref{eq:an_function_mapping_minus_1_1} below.

@@

The \acf{ReLU} \index{activation function}activation function is an \index{activation function}activation function defined in the positive part of its argument and is given in Equation \eqref{eq:relu} below.

@@

\noindent
An advantage of \acs{ReLU} is that it is not susceptible to the \index{vanishing gradient problem}vanishing gradients problem \cite{ref:xu:2015, ref:maksutov:2018} which occurs when gradients in the first layers of a multi-layer \acs{ANN} approach zero, and have no effect on the training process.

@@

The \acf{LReLU} \index{activation function}activation function is a variant of the \acs{ReLU} \index{activation function}activation function that avoids zero gradients in the negative part of its argument by introducing a scaling parameter, $\alpha > 0$ \cite{ref:xu:2015}. Similar to \acs{ReLU}, \acs{LReLU} is not susceptible to the \index{vanishing gradients problem}vanishing gradients problem. However, \acs{LReLU} does not suffer from the \index{dying ReLU problem}dying \acs{ReLU} problem \cite{ref:agarap:2018}, which occurs when the neurons that use \acs{ReLU} \index{activation functions}activation functions become \textit{inactive} and only output $0$ due to a negative net input signal. The \acs{LReLU} \index{activation function}activation function is defined in Equation \eqref{eq:leaky_relu} below.

@@

\noindent
By introducing a parameter $\alpha > 0$, negative \index{Net Input Signals}net input signals will still yield non-zero activations, resulting in non-zero gradients and avoiding gradient saturation. Non-zero gradients are required by some heuristics such as \ac{SGD} in order to be able to effectively train \acp{ANN} \cite{ref:hanin:2018}. An illustration of \acs{LReLU} with various values for $\alpha$ and $\theta = 0$ is given in Figure \ref{fig:anns:activation_functions:leaky_relu}.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\textwidth]{lrelu.pdf}
    \caption[The \ac{LReLU} \index{activation function}activation function]{An illustration of the \acs{LReLU} \index{activation function}activation function with various values for $\alpha$ and $\theta = 0$.}
    \label{fig:anns:activation_functions:leaky_relu}
\end{figure}

@@

The \index{sigmoid}sigmoid activation is the continuous differentiable approximation of the step function, which was used in the original perceptron model developed by \citeauthor{ref:rosenblatt:1957} \cite{ref:rosenblatt:1957}, and yields output in the range $(0, 1)$. The \index{sigmoid}sigmoid \index{activation function}activation function is defined in Equation \eqref{eq:sigmoid} below.

@@

\noindent
In Equation \eqref{eq:sigmoid} above, $\lambda$ is a control parameter, which controls the steepness of the sigmoid activation function and is usually set to $\lambda = 1$. An illustration of the \index{sigmoid}sigmoid \index{Activation Function}activation function with various values for $\lambda$ and $\theta = 0$ is given in Figure \ref{fig:anns:activation_functions:sigmoid}.

@@

The \index{hyperbolic tangent}hyperbolic tangent \index{activation function}activation function has a similar shape to that of the \index{sigmoid}sigmoid \index{activation function}activation function, but yields output in the range $(-1, 1)$. The \index{hyperbolic tangent}hyperbolic tangent \index{activation function}activation function is given in Equation \eqref{eq:hyperbolic_tangent} below.

@@

Output signals are numerical values that represent the output of the \acs{AN}'s \index{activation function}activation function. Output signals are often referred to as \textit{predictions}. Output values need to be post-processed to ensure the practical application of the \ac{AN}.

@@

The \index{softmax}softmax function, also known as the \index{softargmax}\textit{softargmax} \cite[p.~184]{ref:goodfellow:2016} or \index{normalised exponential function}\textit{normalised exponential function} \cite{ref:bishop:2006} is a generalisation of the logistic function that converts a $K$-dimensional output vector $\vec{y}$ into a $K$-dimensional output vector $\vec{y^{'}}$ where each element $y^{'}_k$ is in the range $(0,1)$ and all elements sum up to $1$ as is shown in Equation \eqref{eq:sum_after_softmax} below.

@@

\noindent
The softmax function is then given in Equation \eqref{eq:softmax} below.

@@

The \index{Argmax}argmax function is similar to the \index{softmax}softmax function, with the difference that the element $y_k, k \in \{1,2, \dots, K\}$ with the highest output value is set to $1$ and the rest are set to $0$. All elements still sum to $1$, but the activation is only observed at index $k$ where the activation is $1$. The \index{Argmax}argmax function is given in Equation \eqref{eq:argmax} below.

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.75\textwidth]{output_modifiers.pdf}
    \caption[The results of \index{softmax}softmax and \index{Argmax}argmax]{An illustration of \index{softmax}softmax and \index{Argmax}argmax applied after the \index{sigmoid}sigmoid \index{activation function}activation function.}
    \label{fig:anns:activation_functions:softmax_argmax}
\end{figure}

@@

\noindent
The composition of \acp{AN} in an \ac{ANN} is expressed as the \textit{architecture} and \textit{topology} of the \ac{AN}. The exact composition to use depends on the problem being solved.

@@

The input layer contains the input data to the \ac{ANN}. Since the input layer simply provides the input data, some literature do not consider the input layer as an actual part of the \ac{ANN} \cite{ref:engelbrecht:2007}.

@@

\noindent \ac{BinXE} is used in classification problems, where there are only
two classes in the target output data.

>>>>

\noindent
The human brain is one of the most complicated biological organs in nature. It gives us the ability to think and learn. The field of \acs{ML} has taken a lot of inspiration from the biological brain which lead to the development of \acp{ANN} \cite{ref:rosenblatt:1958}. \acp{ANN} are used throughout this dissertation as the \textit{model} to be trained using \acp{HH}.

The purpose of this chapter is to provide the necessary background information needed on \acp{ANN} and is structured as follows.

\begin{itemize}
    \item \textbf{Section \ref{sec:anns:bn}} gives background information on the \acs{BN}.

    \item \textbf{Section \ref{sec:anns:an}} introduces the \acs{AN}. Brief discussions follow on input, \index{weights}weights and \index{biases}biases, \index{net input signal}net input signal, \index{activation function}activation functions, and output.

    \item \textbf{Section \ref{sec:anns:ann}} introduces the \acs{ANN}. Brief discussions follow on \acs{ANN} architecture, topology, and \acp{FFNN}.

    \item \textbf{Section \ref{sec:anns:training}} provides details on the training process, \index {supervised learning}supervised learning, training sets, stopping conditions, performance measures and \index{error function}error functions.

@@

\index{biological neuron}
\section{Biological Neuron}\label{sec:anns:bn}

This section introduces the \acs{BN} and provides the necessary background information that show how the \acs{BN} has inspired the development of the \acs{AN}.

@@

\noindent
The biological neural systems are made up of microscopic nerve cells called neurons \cite{ref:jain:1996}. Figure \ref{fig:biological_neuron} illustrates a single \acs{BN}. The main components of the \acs{BN} include the cell body, \index{dendrites}\textit{dendrites} and the \index{axon}\textit{axon}. \Acp{NN} are formed through the connections between the \index{axon}axons and \index{dendrites}dendrites of various neurons. This is known as \index{synaptogenesis}\textit{synaptogenesis} \cite{ref:huttenlocher:1997}. Such a connection is referred to as a \index{synapse}\textit{synapse}. Communication takes place, through the synapse, by electro-chemical pulse and is often referred to as an \textit{activation} or \textit{action potential}.  Communication signals propagate from the \index{dendrites}dendrites, through the cell body to the axon of a neuron, provoking a signal in the post-synaptic neuron \cite{ref:engelbrecht:2007}. The greater the connection between two neurons, the stronger the communication.  Kennedy \cite{ref:kennedy:2016} defines stronger \index{synapse}synapses as ones that contribute more depolarisation to the neural membrane upon activation than weaker ones. Stronger \index{synapse}synapses have a higher probability of generating an action potential in the post-synaptic neuron. During activation, the pre-synaptic neuron release neurotransmitters that bind to the post-synaptic neuron \cite{ref:khanacademy:synapse}. The frequent release of these molecules cause the \index{synapse}synapse to grow. Connections that grow over time yield stronger signals (learning), while connections that are weak propagate low intensity signals and vanish over time (forgetting). The ability of synapses to strengthen and weaken over time is known as \index{synaptic plasticity}\textit{synaptic plasticity} \cite{ref:huttenlocher:1997}.

\index{artificial neuron}
\section{Artificial Neuron}\label{sec:anns:an}

This section introduces the \acs{AN}. Brief discussion follow on the various components that make up the \acs{AN}.

The \acs{AN} implements a nonlinear mapping from $\mathbb{R}^{I}$ to
$\mathbb{R}^{T}$, usually in the ranges $[0,1]$ or $[-1,1]$, depending on the
activation function used \cite{ref:engelbrecht:2007} and is given
in Equation~\eqref{eq:an_function_mapping} below.

@@

\noindent
In Equation~\eqref{eq:an_function_mapping} above, $f_{AN}$ is the mapping function produced by the \acs{AN}, $I$ is the total number of dimensions of the input in real-number space ($\mathbb{R}$), and $T$ is the total number of dimensions of the target (desired output) in real-number space.

The \acs{AN} implements various components and is illustrated in Figure \ref{fig:artificial_neuron}.

@@

\noindent
Each of the components of the \acs{AN} is inspired by some element of the \acs{BN}. Brief descriptions of these are given as follows.

@@

    \item \textbf{Biases:} The biases model a mechanism introduced to influence the strength of the activation (output signal) of the \acs{AN}. The bias is represented by $\theta$, in Figure \ref{fig:artificial_neuron}.

    \item \index{activation function}\textbf{Activation function:} The activation function models the action potential of the \acs{BN} and is based on the net input signal. The \index{activation function}activation function is represented by $f$, in Figure \ref{fig:artificial_neuron}.

@@

Input signals are $I$-dimensional vectors of numerical values that are obtained either through some environment sensor or from other \acp{AN}. Input signals are often referred to as \index{features}\textit{features} or \index{independent variables}\textit{independent variables} \cite{ref:francis:2001}. Throughout this dissertation, a single input vector is referred to as a \textit{pattern}. Input data must first be pre-processed before it is presented to the \acs{AN}. Input pre-processing techniques are presented in the following sections.

@@

Input pre-processing improves the training process and contributes to the success of the practical application of the \acs{ANN} \cite{ref:kuzniar:2017}. The primary purpose of input pre-processing is to modify the input data so that it can better match predicted output data. There are many techniques to consider when pre-processing input data. These techniques include methods to encode data to certain formats, to scale to the correct ranges and to deal with incomplete, invalid or irrelevant data. For the purposes of this dissertation, \textit{data encoding} and \textit{normalisation} techniques are considered.

\index{encoding}
\subsubsection{Encoding}\label{sec:anns:an:input:encoding}

For qualitative data, the class labels have to be converted from textual to numerical representations. In general, class labels are encoded as numerical vectors \cite{ref:srinidhi:2018, ref:brownlee:2017:one-hot}. One such encoding technique is referred to as \index{one-hot encoding}\textit{one-hot encoding}. Harris and Harris \cite{ref:harris:2010} describe \index{one-hot encoding}one-hot encoding as a group of vector elements where the classes are represented with a single high element (represented by $1$) and all the others low (represented by $0$). Each element $y_k$, where $k$ is the $k$-th element of the encoded vector $\vec{y}$, uniquely refers to a class. An activation, represented by a value of $1$ at a given index thus represents a class associated with that index.

\index{normalisation}
\subsubsection{Normalisation}\label{sec:anns:an:input:normalisation}

Numerical data must be normalised and scaled appropriately for the \acs{AN}. In general, input data is normalised to a range that is appropriate for the \index{activation function}activation function being used by the \acs{AN}. Normalisation and scaling yields input data that is comparable to the output produced by the \acs{AN}. For the purposes of this disseration, the \index{min-max scaler}min-max scaler \cite{ref:al:2006} and the \index{standard score scaler}standard score scaler \cite{ref:jain:2005} are considered.

\index{min-max scaler}
\subsubsection{Min-Max Scaler}\label{sec:anns:an:input:min_max_scaler}

The \index{min-max scaler}min-max scaler, also called \index{unity-based normalisation}\textit{unity-based normalisation}, scales values to the range $[0,1]$. The \index{min-max scaler}min-max scaler is used as a pre-processing technique for target values when the \index{sigmoid}\textit{sigmoid} \index{activation function} activation function is used. The \index{min-max scaler}min-max scaler is given in Equation~\eqref{eq:min_max_scaler} below.

@@

\noindent
In Equation~\eqref{eq:min_max_scaler} above, $x_{i,p}^{'}$ is the normalised form of $x_{i,p}$, the $i$-th dimension of the input vector $\vec{x}_p$, $p \in \{1,2, \dots, P \}$, where $P$ is the total number of input patterns, $x_{i_{min}}$ and $x_{i_{max}}$ are respectfully the minimum and maximum values of the $i$-th dimension for all input vectors $\vec{x}_p$.

\index{standard score scaler}
\subsubsection{Standard Score Scaler}\label{sec:anns:an:input:standard_score_scaler}

The \index{standard score scaler}standard score scaler, also known as the \index{z-score scaler}\textit{z-score scaler}, scales values by subtracting the mean and scaling to the unit variance of each dimension $i$ for all input patterns $\vec{x}_p, p \in \{1,2, \dots, P\}$. The \index{standard score scaler}standard score scaler is used as a pre-processing technique for target values when the \index{hyperbolic tangent}\textit{hyperbolic tangent} \index{activation function}activation function\index{activation function} is used. The \index{standard score scaler}standard score scaler is given in Equation~\eqref{eq:standard_score_scaler} below.

@@

\noindent
In Equation~\eqref{eq:standard_score_scaler} above, $\mu_i$ and $\sigma^2_i$ are respectfully the mean and unit variance of the $i$-th dimension of all input vectors $\vec{x}_p$.

@@

The connection strength that synapses in the \acs{BN} have is modeled in the \acs{AN} as weight vectors of numerical values associated with each dimension of the input. Weights can either dampen or strengthen the input by some negative or positive numerical value. Changes in the weight associated with a feature changes the influence that that particular feature has on the predicted output. Finding the correct value for each weight, such that the \acs{AN} yields optimal output, is an optimisation problem \cite{ref:thierens:1993}. Research has shown that weight initialisation plays an important role in the efficient training of \acp{ANN} \cite{ref:thimm:1995}.

\subsubsection{Weight Initialisation}\label{sec:anns:an:weights:initialisation}

Weight initialisation is the process by which candidate solutions (represented by the weights of the \acs{AN}) to the problem are ``placed'' in the search-space. Weight initialisation influences the speed of convergence, the probability of convergence and the generalisation capabilities of \acp{ANN} \cite{ref:fernandez:2001}. Finding the optimal initialisation values for weights is non-trivial and can be seen as another optimisation problem \cite{ref:de:2016, ref:erdogmus:2003, ref:yam:2000}.

Weight initialisation is dependent on the \index{activation function}activation function used. If weights are initialised as values that are too small, the vanishing gradients problem can occur \cite{ref:hanin:2018}. If weights are initialised as values that are too big, output of the activation function would not be in the active range. This leads to unit saturation, and exploding gradients can occur \cite{ref:hanin:2018, ref:yadav:2018}.

@@

\index{Glorot uniform sampling}Glorot uniform sampling is a specialisation of \index{random uniform sampling}random uniform sampling, whereby $\omega = \sqrt{\frac{6}{fanin + fanout}}$ and $fanin$ is the number of input neurons to the weight vector and $fanout$ is the number of output neurons from the weight vector.

\index{Glorot normal sampling}
\subsubsection{Glorot Normal Sampling}\label{sec:anns:an:weights:glorot_normal_sampling}

@@

\acp{AN} accumulate the net resulting input signal from all input dimensions into a value called the \index{net input signal}\textit{net input signal}, expressed as $net$. This signal is passed to the \index{activation function}activation function, which provokes an artificial action potential in the \acs{AN}.

A common way by which the \index{net input signal}net input signal is calculated is by means of \acp{SU} \cite{ref:engelbrecht:2007}, which compute the \index{net input signal}net input signal as the weighted sum of all input signals as shown in Equation~\eqref{eq:summation_units} below.

@@

\noindent
In Equation~\eqref{eq:summation_units} above, $x_{i}$ is the $i$-th dimension of the input vector $\vec{x}$, and $v_{i}$ is the $i$-th dimension of the weight vector $\vec{v}$, associated with the given input dimension.

@@

A bias/threshold term $\theta$ is introduced to help translate the output of the \index{activation function}activation function \cite{ref:benitez:1997}. The value of $\theta$ can be learned during the training process along with the weights of the \acs{ANN}. In order to simplify equations, the input and weight vectors are augmented such that the input and hidden layers have an additional neuron/unit, called the \textit{bias unit} \cite{ref:engelbrecht:2007}. The net input signal can then be rewritten to consider the bias unit, leading to an augmented net input signal.

@@

The augmented net input signal, that includes the bias unit, has the form $net^{'} = net - \theta$, with $\theta = x_{i+1}v_{i+1}$. A constant value $x_{i+1} = -1$ can be used, meaning that the weight associated with the bias unit, $v_{i+1}$, is optimised along with the other weights during the optimisation process. The net input signal as given in Equation~\eqref{eq:summation_units} then changes as is shown in Equation~\eqref{eq:augmented_vectors} below.

@@

An \index{activation function}activation function is used to model the action potential of the \acs{AN} \cite{ref:ziv:1994, ref:hodgkin:1952}. The \index{activation function}activation function takes as a parameter, the augmented \index{net input signal}net input signal. When the output produced by the activation function surpasses some threshold value $\tau$, we consider that neuron to have ``fired'' an output signal. \index{activation function}Activation functions thus model \textit{phase shift}. In the context of classification problems, \index{activation function}activation functions form decision boundaries between classes. In the context of regression problems, \index{activation function}activation functions try to approximate some function that maps the input data to some target data.

In general, activation functions produce a non-linear mapping of $\mathbb{R}^{I}$ to the range $[0,1]$ or $[-1,1]$ as shown in Equations~\eqref{eq:an_function_mapping_0_1} and~\eqref{eq:an_function_mapping_minus_1_1} below.

@@

The \acf{ReLU} \index{activation function}activation function is an \index{activation function}activation function defined in the positive part of its argument and is given in Equation~\eqref{eq:relu} below.

@@

\noindent
An advantage of \acs{ReLU} is that it is not susceptible to the \index{vanishing gradients problem}vanishing gradients problem \cite{ref:xu:2015, ref:maksutov:2018} which occurs when gradients in the first layers of a multi-layer \acs{ANN} approach zero, and have no effect on the training process.

@@

The \acf{LReLU} \index{activation function}activation function is a variant of the \acs{ReLU} \index{activation function}activation function that avoids zero gradients in the negative part of its argument by introducing a scaling parameter, $\alpha > 0$ \cite{ref:xu:2015}. Similar to \acs{ReLU}, \acs{LReLU} is not susceptible to the \index{vanishing gradients problem}vanishing gradients problem. However, \acs{LReLU} does not suffer from the \index{dying ReLU problem}dying \acs{ReLU} problem \cite{ref:agarap:2018}, which occurs when the neurons that use \acs{ReLU} \index{activation function}activation functions become \textit{inactive} and only output $0$ due to a negative net input signal. The \acs{LReLU} \index{activation function}activation function is defined in Equation~\eqref{eq:leaky_relu} below.

@@

\noindent
By introducing a parameter $\alpha > 0$, negative \index{net input signal}net input signals will still yield non-zero activations, resulting in non-zero gradients and avoiding gradient saturation. Non-zero gradients are required by some heuristics such as \acs{SGD} in order to be able to effectively train \acp{ANN} \cite{ref:hanin:2018}. An illustration of \acs{LReLU} with various values for $\alpha$ and $\theta = 0$ is given in Figure \ref{fig:anns:activation_functions:leaky_relu}.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\textwidth]{lrelu.pdf}
    \caption[The \acs{LReLU} \index{activation function}activation function]{An illustration of the \acs{LReLU} \index{activation function}activation function with various values for $\alpha$ and $\theta = 0$.}
    \label{fig:anns:activation_functions:leaky_relu}
\end{figure}

@@

The \index{sigmoid}sigmoid activation is the continuous differentiable approximation of the step function, which was used in the original perceptron model developed by \citeauthor{ref:rosenblatt:1957} \cite{ref:rosenblatt:1957}, and yields output in the range $(0, 1)$. The \index{sigmoid}sigmoid \index{activation function}activation function is defined in Equation~\eqref{eq:sigmoid} below.

@@

\noindent
In Equation~\eqref{eq:sigmoid} above, $\lambda$ is a control parameter, which controls the steepness of the sigmoid activation function and is usually set to $\lambda = 1$. An illustration of the \index{sigmoid}sigmoid \index{activation function}activation function with various values for $\lambda$ and $\theta = 0$ is given in Figure \ref{fig:anns:activation_functions:sigmoid}.

@@

The \index{hyperbolic tangent}hyperbolic tangent \index{activation function}activation function has a similar shape to that of the \index{sigmoid}sigmoid \index{activation function}activation function, but yields output in the range $(-1, 1)$. The \index{hyperbolic tangent}hyperbolic tangent \index{activation function}activation function is given in Equation~\eqref{eq:hyperbolic_tangent} below.

@@

Output signals are numerical values that represent the output of the \acs{AN}'s \index{activation function}activation function. Output signals are often referred to as \textit{predictions}. Output values need to be post-processed to ensure the practical application of the \acs{AN}.

@@

The \index{softmax}softmax function, also known as the \index{softargmax}\textit{softargmax} \cite[p.~184]{ref:goodfellow:2016} or \index{normalised exponential function}\textit{normalised exponential function} \cite{ref:bishop:2006} is a generalisation of the logistic function that converts a $K$-dimensional output vector $\vec{y}$ into a $K$-dimensional output vector $\vec{y^{'}}$ where each element $y^{'}_k$ is in the range $(0,1)$ and all elements sum up to $1$ as is shown in Equation~\eqref{eq:sum_after_softmax} below.

@@

\noindent
The softmax function is then given in Equation~\eqref{eq:softmax} below.

@@

The \index{argmax}argmax function is similar to the \index{softmax}softmax function, with the difference that the element $y_k, k \in \{1,2, \dots, K\}$ with the highest output value is set to $1$ and the rest are set to $0$. All elements still sum to $1$, but the activation is only observed at index $k$ where the activation is $1$. The \index{argmax}argmax function is given in Equation~\eqref{eq:argmax} below.

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.75\textwidth]{output_modifiers.pdf}
    \caption[The results of \index{softmax}softmax and \index{argmax}argmax]{An illustration of \index{softmax}softmax and \index{argmax}argmax applied after the \index{sigmoid}sigmoid \index{activation function}activation function.}
    \label{fig:anns:activation_functions:softmax_argmax}
\end{figure}

@@

\noindent
The composition of \acp{AN} in an \acs{ANN} is expressed as the \textit{architecture} and \textit{topology} of the \acs{AN}. The exact composition to use depends on the problem being solved.

@@

The input layer contains the input data to the \acs{ANN}. Since the input layer simply provides the input data, some literature do not consider the input layer as an actual part of the \acs{ANN} \cite{ref:engelbrecht:2007}.

@@

\noindent
